Tropical Mathematics for the Lambda Calculus

### 2023-01-19 section4.tex
commit
Commit

A tropical polynomial, as we saw, is a piece-wise linear function of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$. For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
%\begin{proof}
%Call $\leq$ the well-founded pointwise order on $\N^k$.
%Now we can let $\C F_\epsilon:=\set{n\in\N \mid
%\widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
%\end{proof}

\subsection{Continuity of tLs}%$\Lawv^{X}$ as a normed cone}

$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This order makes it a Scott-continuous dcpo.
Furthermore we have:

\begin{proposition}
  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
  % on the domain and codomain.
\end{proposition}
The tLs $\varphi$ is continuous on $\BB R_{\geq0}=\Lawv-\set{\infty}$ (w.r.t.\ the usual norm of real numbers).
By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we could generalise this property by dropping the case of $x$ having some $0$ coordinate:

\begin{theorem}\label{thm:cont}
 Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
\end{theorem}
\begin{proof}
 The result follows after proving that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
\end{proof}

@@

\begin{theorem}\label{thmTLSlocLip}
 All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}$.
\end{theorem}
\begin{proof}[Proof sketch]
The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz. 
\end{proof}

>>>>

A tropical polynomial, as we saw, is a piece-wise linear function of the form $f(x)=\min_{i_{1},\dots, i_{k}}\{i_{j}x+c_{i_{j}}\}$. For instance, the polynomials $\varphi_{n}(x)=\min_{i\leq n}\{in+2^{-i}\}$
are plotted in Fig.~\ref{fig:plot1} for $n=2,3,4$.
A \emph{tropical root} of a tropical polynomials $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable. In other words, the roots of $\varphi$ are the points where the minimum defining $\varphi$ is attained at least twice (i.e.~where the slope of $\varphi$ changes. For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $i \leq n$.
With this definition, tropical roots mimic the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as
$f(x)=\min\{x,x_{0}\}+ g(x)$. Yet, unlike in standard algebra, tropical roots can be computed in linear time \cite{Noferini2015}.
%
%A
% simple calculation shows that this condition corresponds, in the tropical setting, to the usual notion of root.

@@

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\matr f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with the tropical {polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}
Call $\leq$ the (well-founded) pointwise order on $\N^k$.
Now \mbox{$\C F_\epsilon:=\set{n\in\N \mid
\widehat f(n)<\infty \textit{ and } 
\widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$}
does the job.
\end{proof}

\subsection{Continuity of tLs}\label{subsec:cont}%$\Lawv^{X}$ as a normed cone}

The tLs $\varphi$ is continuous on $\BB R_{\geq0}$ (w.r.t.\ the usual norm of real numbers).
By considering the usual norm $\norm{x}_\infty:=\sup\limits_{a\in X} \absv{x_a}$ on $\Lawv^X$, we can generalise this property by dropping the case of $x$ having some $0$ coordinate:

\begin{theorem}\label{thm:cont}
 Any tropical Laurent series $f:\Lawv^X\to\Lawv$ is continuous on $\BB R_{>0}^X$, w.r.t.\ to the norm $\norm{\cdot}_\infty$.
\end{theorem}
\begin{proof}
 It follows after adaptating [Theorem , \cite{}] in order to prove that if a real-valued function on a locally convex topological $\BB R$-vector space is, locally around $x$, concave and bounded by a finite constant, then it is continuous at $x$.
\end{proof}

We conclude this subsection by noticing that $\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete \emph{normed cone} %(see~\cite{Selinger2004}, or the appendix, for such notions}.
Its cone structure induces an order on it, called its \emph{cone order}:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This makes it a Scott-continuous dcpo.
Suitable categories of cones have been recently investigated as models of probabilistic computation %(\cite{Crubillie2018, EhrPagTas2018, Ehrhard2020}).
We will just show that:

\begin{theorem}\label{thm:ScottCont}
  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders.
  % on the domain and codomain.
\end{theorem}

@@

\begin{theorem}\label{thmTLSlocLip}
 All tropical Laurent series $\Lawv^X\to\Lawv$ are locally Lipschitz on $\BB R_{>0}^X$.
\end{theorem}
\begin{proof}[Proof sketch]
The core of the proof is a convex analysis argument (see the Appendix) showing that an arbitrary function $f:\Lawv^X\to\Lawv$ which is non-decreasing, concave and continuous, must be locally Lipschitz. 
\end{proof}

### 2023-01-19 section5.tex
Commit

Let us consider in this subsection a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
\[
 \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
 \qquad\qquad
 \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
\]
We add reduction rules:
\[
 M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
\]
so that $M\oplus_p N$ plays the role of a probabilistic sample with parameter $p$.
We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$, starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

This way the probabilistic evaluation of $M$ is presented in the form of a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}. Indeed, typical questions in this case would be:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
 This means asking which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ by repeatedly sampling $M$.
 \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e. that among the paths leading to $\true$, the one taken was $\omega_0$)?
\end{enumerate}
A similar argument could be done by replacing $\true$ and $\false$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}$, which is more easily solved by 
passing to the tropical polynomial $\trop P_{\omega}$. 
For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
\begin{equation}
  \begin{array}{r@{}c@{}l@{}}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   & = &
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   \\
   & = &
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
  \end{array}
\end{equation}
where the equalities hold for all $0<c\neq1$.

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing to $\trop P_{\omega}$ makes the problem easier, as this may be reduced to computing the tropical roots of the polynomial (which can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

\begin{remark}\label{rem:troproots}
 Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
 Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.
\end{remark}

>>>>

Let us consider in this subsection a probabilistic extension of $\STLC$, call it $\STLC_\oplus$:
we add a ground type $\bool$, terms $\true,\false$ of type $\bool$, terms of shape $M\oplus_p N$ and $pM$, for $p\in[0,1]$, typed via the rules:
\[
 \dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A} 
 \qquad\qquad
 \dfrac{\Gamma\vdash M:A}{\Gamma\vdash pM:A}
\]
We add reduction rules:
\[
 M\oplus_p N \to pM \quad \textit{ and } \quad M\oplus_p N \to (1-p)N
\]
so that $M\oplus_p N$ plays the role of a probabilistic coin toss of bias $p$.
We use this calculus as a toy example for our purposes, and it can be seen as a fragment of the PCF considered in \cite{Manzo2013}.

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$ starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ after all the coin tossings in $M$ are performed.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after all the tossings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

@@

This way, the probabilistic evaluation of $M$ is presented as a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}.
Typical questions in this case would be:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
 I.e., which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ after all tossings in $M$?
 \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e. that among the paths leading to $\true$, the one taken was $\omega_0$)?
\end{enumerate}
A similar argument could be done by replacing $\true$ and $\false$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}, Q_\omega$, which is more easily solved by 
passing to the tropical monomials/polynomial $\trop P_{\omega},\trop Q_{\omega}$. 
For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
\begin{equation}
  \begin{array}{r@{}c@{}l@{}}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   & = &
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   \\
   & = &
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
  \end{array}
\end{equation}
where the equalities hold for all $0<c\neq1$.

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing through $\trop$ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

\begin{remark}\label{rem:troproots}
 Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the bias of $\oplus_p$ for which there are at least two different equiprobable paths of tossings from $M$ to its normal form.
 Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the bias of $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are smapled by $M$ during its tossings to normal form, knowing that some $\true$ was sampled.
\end{remark}
